KIN: A method to infer relatedness from low-coverage ancient DNA

### 2022-09-11 paper_overleaf/main.tex

\noindent Inferring genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses. However, estimating relatedness from ancient DNA is difficult due to low-coverage, ascertainment bias, or contamination from various sources. Here, we present KIN, a method to estimate the relatedness of a pair of individuals from the identical-by-descent segments they share. KIN accurately classifies up to 3rd-degree relatives using >0.05x sequence coverage and differentiates siblings from parent-child. It incorporates additional models to adjust for contamination and detect inbreeding, which improves classification accuracy, and can be applied to any diploid species. 
\end{abstract}

@@

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), i.e. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, each parent will share exactly one set of chromosomes IBD with their offspring, while subsequent recombination means that a grandparent will, on average share a quarter of their genome with a grand-child. Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: the individuals share zero, one or two chromosomes IBD. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig. \ref{fig0:schematic}). These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD. In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.

@@

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case (section \ref{simulat}). The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to the stationary distribution of the transition matrix for simplicity. The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

@@

\begin{align}\label{eq:1}
\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
&= \sum_w \log P(D_w|Z_w,H_w, N_w, \delta) + \sum_w \log P(Z_w |Z_{w-1},A) + \log P(Z_0|\pi)\text{.}
\end{align}

Here, $\BZ$ is not dependent on $\BH$ and $\delta$.

@@

Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them for $\delta$. The simplest model is to assume that sites in each window are equally distributed and independent, in this case we could use the  binomial likelihood:
$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$

@@

Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
\begin{equation}\label{eq:2}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_0 & p_0 \\
p_1 & p_2 & p_4 \\
p_2 & p_4 & p_4
\end{array}\right]
\end{equation}

@@

\paragraph{Beta Binomial Model:}
We empirically find that the data often has considerably higher variance than would be expected from a binomial model (Fig. S2). We take this into account by adding an over-dispersion parameter $\delta$. Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter.

Taken together, our emission probabilities are 
\begin{align}\label{eq:3}
P(D_{w}|Z_w,H_w,N_w, \delta) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
\end{align}
where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see equation \ref{eq:2}).

@@

We constrain the optimization space of the $\delta_i$, as unconstrained optimization could result in some confounding of cases. We know that different cases of relatedness have different numbers of IBD states possible. For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$. However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance (for example, see Fig. S5, S6). We avoid this problem by constraining the $\delta$ such that the beta distributions for the different relatedness cases overlap by at most one standard deviation.

@@

\subsection{ROH estimation model}\label{roh}
Our HMM to detect ROH tracts works similarly to the KIN-HMM described above, but in this case we only consider one individual at a time, and only consider positions covered by at least two reads. For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome. We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{M}$. Our model has two possible hidden states: homozygous state ($Y_w=3$), and non-homozygous state ($Y_w=2$). As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$. The complete data likelihood for the model in this case is then:

\begin{align}\label{eq:10}
    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
 &= [\sum_{w} log P(\Delta_w|Y_w, M_w, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, A)] + P(Y_0| \pi),
\end{align}

where $\Theta$ is a vector of initial probabilities ($\pi$), transition matrix ($A$) and $M$.

Since the source of ROH may not be known, we estimate both transitions and emissions. We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 3 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_4$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively. The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window. The maximization step for emissions is analogous to that in the KIN model (eq. \ref{eq:4}), and the optimization step here is done with the following cost function:

\begin{align}\label{eq:11}
\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
\end{align}
where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq. \ref{eq:3}.

@@

\subsection{Evaluation with simulations}
We first tested the performance of KIN with simulated pedigrees. We performed coalescent simulations to generate 8 unrelated diploid genomes and artificially mated them to form pedigrees of 17 individuals with relationships up to $5^{th}$ degree (Fig. S3). To evaluate the effect of sequence coverage on the performance of KIN, we generated artificial pileups at each polymorphic site for each individual following a Poisson distribution with 6 different average depths varying between 4$x$ and 0.03$x$ (see section \ref{method}). To mimic ROH, for some pedigrees, we picked a single allele at heterozygous sites in some regions as determined by a Markov chain so that about $17\%$ of the genome is ROH. We also created versions with contamination, by introducing alleles from distantly related individuals, and ascertainment bias, by selecting polymorphic sites identified in a subset of the individuals (see section \ref{method}). We created 60 pedigrees for each combination of average coverage and scenarios of presence/absence of ROH, contamination, and ascertainment bias, totalling 2,880 pedigrees.

\paragraph{ROH detection:}
We first evaluated the performance of the ROH-HMM ( Fig.\ref{fig2:ROH}). We computed the proportion of ROH in all windows and compared it to the posterior probability of the ROH state, by computing the bias (i.e., the difference between both) and its variance across windows. Without ROH, we correctly inferred the absence of ROH as the observed bias ranges from $0.3$ to $0.7\%$ (variance $\leq0.006$; Table \ref{tab:Table 2}). With ROH, we observed a slight bias of $5$ to $9\%$ more ROH than simulated (variance $\leq0.04$). The largest bias observed ($10.1\%$) was for the datasets that included contamination. Ascertainment bias had a minimal effect on the detection of ROH (Table \ref{tab:Table 2}). Surprisingly, the bias was lower for lower coverages, perhaps because there is only one possible ROH state per window to model a continuous variable, viz. proportion of window in ROH, and this state is inferred more confidently with more data. Overall, the model predicts ROH well in all cases, even at a coverage of 0.1x (Fig.\ref{fig2:ROH}).

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_review.png}
    \centering
    \caption{Estimation of ROH probabilities along the genome in simulations. Top panel shows proportion of differences in a simulated individual in windows along the genome, and the bottom panel shows probability of seeing no ROH. (A) Simulation with no ascertainment, contamination, or ROH. (B) Simulation with ROH. (C) Simulation with ROH and ascertainment. (D) Simulation with ROH and contamination.}
    \label{fig2:ROH}
\end{figure}

\begin{table}[h!]
\caption{\label{tab:Table 2}Model performance for ROH prediction. Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
    \hline
    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RA(bias) & RA(var) & RC(bias) & RC(var)\\
    \hline
    4x & 0.007 & 0.006 & 0.087 & 0.038 & 0.089 & 0.039 & 0.101 & 0.042\\
    \hline
    0.5x & 0.006 & 0.005 & 0.085 & 0.036 & 0.087 & 0.036 & 0.098 & 0.039\\
    \hline
    0.2x & 0.004 & 0.003 & 0.073 & 0.03 & 0.073 & 0.028 & 0.084 & 0.031\\
    \hline
    0.1x & 0.003 & 0.002 & 0.054 & 0.023 & 0.054 & 0.023 & 0.065 & 0.025\\
    \hline
\end{tabular}
\label{table1}
\end{table}

\begin{table}[h!]
\caption{\label{tab:Table 3}Model performance for ROH prediction. Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
    \hline
    Coverage & R(Sensitivity) & R(Specificity) & RA(Sensitivity) & RA(Specificity) & RC(Sensitivity) & RC(Specificity)\\
    \hline
    4x & 0.007 & 0.006 & 0.087 & 0.038 & 0.089 & 0.039 & 0.101 & 0.042\\
    \hline
    0.5x & 0.006 & 0.005 & 0.085 & 0.036 & 0.087 & 0.036 & 0.098 & 0.039\\
    \hline
    0.2x & 0.004 & 0.003 & 0.073 & 0.03 & 0.073 & 0.028 & 0.084 & 0.031\\
    \hline
    0.1x & 0.003 & 0.002 & 0.054 & 0.023 & 0.054 & 0.023 & 0.065 & 0.025\\
    \hline
\end{tabular}
\label{table2}
\end{table}

\paragraph{IBD prediction:}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatedness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.78, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.48 at 0.03x. The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage. We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction. Adding ROH to the simulations reduces the IBD prediction in two cases: the accuracy for second degree is decreased by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig. S9). We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}). Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$). However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see Fig. \ref{fig3:Comparison_READ_KIN})

@@

\paragraph{Relatedness classification:}
We evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig. \ref{fig3:Comparison_READ_KIN}). We first describe the results for relatedness cases detectable by READ, viz. identical, $1^{st}$ degree, $2^{nd}$ degree and unrelated. In this case we show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data. The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02. One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.95 to 0.89, compared to range of 0.91 to 0.75 for READ. The false positive rate in this case remains below 0.03 for both methods. We also find that READ is strongly impacted by contamination (proportion of contamination$<3\%$) as the true positive rate is in the range 0.89 to 0.78 and 0.54 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.03 and 0.16 for unrelated individuals and 2nd degree relatives, respectively. By contrast, the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control. For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives. The increase in false positives is likely due to ROH making related individuals more similar. Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e. parent-child, siblings and, to a lower level, 3rd degree relatives (Fig. \ref{fig3:Comparison_READ_KIN}, for details see table S4).

@@

Only for 28 pairs, KIN has confident classifications that differ from those obtained using READ (table S2, Fig. S11). Twenty of these comparisons, KIN predicts to be third-degree relationships, and  READ concordantly classifies them as unrelated (as it only infers first and second degree relationships). lcMLkin predicts $3^{rd}$ - $5^{th}$ in 14 of these cases, unrelated in 1 case (see Fig. S11), second degree in 1 case (see Fig. S11), and does not have enough data for classification in 4 cases. For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^{th}$ degree, but KIN infers them to be unrelated. There are a total of 83 pairs where READ obtains a confident call ($s.d.>1$), but KIN does not ($\Delta LL<1$). In 80 of these cases, READ classifies them as unrelated, KIN classifies $3^{rd}$ degree, while lcMLkin predicts $3^{rd}$ - $5^{th}$ degree. In rest 3 cases READ classifies unrelated, while KIN and lcMLkin agree on $2^{nd}$ degree.

@@

For detecting ROH, there is only one method available that works with low coverage ancient DNA \cite{ringbauer_parental_2021}. This software can infer ROH at coverages $\geq0.3x$, but requires a large reference panel. Instead, our method detects long ROH based on just the expected heterozygosity $p_0$, which can be estimated from a small number of unrelated individuals.  On simulations, we find that our method reliably inferrs ROH regions  with lengths on the order of 10cM from samples sequenced to coverage $\geq 0.1x$ . 
In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages. This is likely because ROH actually reduces the variance in differences depending on the relatedness, and makes it easier to correctly classify relatedness. For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias (see Fig. \ref{fig3:Comparison_READ_KIN}).

We show that KIN reliably detects IBD tracts for up to 0.1x coverage even in presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages. Adding ROH to the simulations adversely affects IBD prediction in siblings (Fig. S9), but this does not affect the power to correctly classify siblings (Fig. \ref{fig3:Comparison_READ_KIN}).

@@

The Lech Valley data has low contamination, and no ROH. For pairwise comparisons with large numbers of overlapping sites ($>10000$), KIN, READ and lcMLkin all mostly agree. However, KIN is able to differentiate between parent-child and siblings, and identify second degree relationship from just a few thousand polymorphic sites ($\approx$ 4000) overlapping between samples. KIN can also infer third degree relation with $\approx$ 30,000 overlapping polymorphic sites. We show that when applied to Neandertal specimens from Chagyrskaya and Okladnikov Caves, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples \cite{laurits_skov_genetic_nodate}. In addition, KIN identifies a pair of $3^{rd}$ degree relatives. In this case of a population with large amounts of ROH, we find that the inference by lcMLkin are heavily biased, but KIN's model takes ROH into account and both the coefficient of relatedness and $k_0$ are very close to what would be expected from the inference by both READ and KIN.

>>>>

\noindent Genetic kinship of ancient individuals can provide insights into their culture and social hierarchy, and is relevant for downstream genetic analyses. However, estimating relatedness from ancient DNA is difficult due to low-coverage, ascertainment bias, or contamination from various sources. Here, we present KIN, a method to estimate the relatedness of a pair of individuals from the identical-by-descent segments they share. KIN accurately classifies up to 3rd-degree relatives using $>0.05x$ sequence coverage and differentiates siblings from parent-child. It incorporates additional models to adjust for contamination and detect inbreeding, which improves classification accuracy. 
\end{abstract}

@@

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), i.e. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, each parent will share exactly one set of chromosomes IBD with their offspring, while subsequent recombination means that a grandparent will, on average share a quarter of their genome with a grand-child. Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: the individuals share zero, one or two chromosomes IBD. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig. \ref{fig0:schematic}). These IBD probabilities can directly be used to categorize their relatedness as shown in Table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD. In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.

@@

There are three additional model parameters, $\bm{\pi}, \mathbf{A}$ and $\bm{\delta}$. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case (section \ref{simulat}). The initial probabilities $\bm{\pi}$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to the stationary distribution of the transition matrix for simplicity. The overdispersion-parameter $\bm{\delta}$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \bm{\pi})$.

@@

\begin{align}\label{eq:1}
\log P(\BD,\BZ|\BH, \theta, \bm{\delta}) &= \log P(\BD|\BZ,\BH, \theta, \bm{\delta}) + \log P(\BZ |\theta) \nonumber\\
&= \sum_w \log P(D_w|Z_w,H_w, N_w, \bm{\delta}) + \sum_w \log P(Z_w |Z_{w-1},\bm{A}) + \log P(Z_0|\bm{\pi})\text{.}
\end{align}

Here, $\BZ$ is not dependent on $\BH$ and $\bm{\delta}$.

@@

Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \bm{\delta})$ from equation \ref{eq:1} and optimize them for $\bm{\delta}$. The simplest model is to assume that sites in each window are equally distributed and independent, in this case we could use the  binomial likelihood:
$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$

@@

Taken together, we can summarize $\bm{p}$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
\begin{equation}\label{eq:2}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_0 & p_0 \\
p_1 & p_2 & p_4 \\
p_2 & p_4 & p_4
\end{array}\right]
\end{equation}

@@

\paragraph{Beta Binomial Model:}
We empirically find that the data often has considerably higher variance than would be expected from a binomial model (Fig. S2). We take this into account by adding an over-dispersion parameter $\bm{\delta}$. Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared, and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter.

Taken together, our emission probabilities are 
\begin{align}\label{eq:3}
P(D_{w}|Z_w,H_w,N_w, \bm{\delta}) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
\end{align}
where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see equation \ref{eq:2}).

@@

We constrain the optimization space of the $\delta_i$, as unconstrained optimization could result in some confounding of cases. We know that different cases of relatedness have different numbers of IBD states possible. For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$. However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance (for example, see Fig. S5, S6). We avoid this problem by constraining the $\bm{\delta}$ such that the beta distributions for the different relatedness cases overlap by at most one standard deviation.

@@

\subsection{ROH estimation model}\label{roh}
Our HMM to detect ROH tracts works similarly to the KIN-HMM described above, but in this case we only consider one individual at a time, and only consider positions covered by at least two reads. For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome. We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{M}$. Our model has two possible hidden states: homozygous state ($Y_w=4$), and non-homozygous state ($Y_w=2$). As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$. The complete data likelihood for the model in this case is then:

\begin{align}\label{eq:10}
    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\bm{\delta}) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \bm{\delta}) + log P(\mathbf{Y}|\Theta)\nonumber\\
 &= [\sum_{w} log P(\Delta_w|Y_w, M_w, \bm{\delta}) + \sum_{w} log P(Y_w|Y_{w-1}, \bm{A})] + P(Y_0| \bm{\pi}),
\end{align}

where $\Theta$ is a vector of initial probabilities ($\bm{\pi}$), transition matrix ($\bm{A}$) and $\bm{M}$.

Since the source of ROH may not be known, we estimate both transitions and emissions. We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 4 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_4$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively. The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=4$ or the state $Y_w=2$ in each window. The maximization step for emissions is analogous to that in the KIN model (eq. \ref{eq:4}), and the optimization step here is done with the following cost function:

\begin{align}\label{eq:11}
\mathcal{C} = \sum_{w=1}^L \sum_{i} P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
\end{align}
where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and over-dispersion parameter $\delta$ similar to eq. \ref{eq:3}, and $i$ can take values 2 and 4 corresponding to the hidden states $Y_w$.

@@

\subsection{Evaluation with simulations}
We first tested the performance of KIN with simulated pedigrees. We performed coalescent simulations to generate 8 unrelated diploid genomes and artificially mated them to form pedigrees of 17 individuals with relationships up to $5^{th}$ degree (Fig. S3). To evaluate the effect of sequence coverage on the performance of KIN, we generated artificial pileups at each polymorphic site for each individual following a Poisson distribution with 6 different average depths varying between 4$x$ and 0.03$x$ (see section \ref{method}). To mimic ROH, for some pedigrees, we picked a single allele at heterozygous sites in some regions as determined by a Markov chain so that on an average about $17\%$ of the genome is ROH. We also created versions with contamination, by introducing alleles from distantly related individuals, and ascertainment bias, by selecting polymorphic sites identified in a subset of the individuals (see section \ref{method}). We created 60 pedigrees for each combination of average coverage and scenarios of presence/absence of ROH, contamination, and ascertainment bias, totalling 2,880 pedigrees.

\paragraph{ROH detection:}
In Fig. \ref{fig2:ROH}, we present an example of data and inference of the ROH-HMM for simulations with and without ROH, potentially with ascertainment bias and contamination. In all cases, we find that the inferred ROH closely matches the simulations, but the confidence in the classification tends to increase with the simulated coverage.

A systematic evaluation of the performance of the ROH-HMM is given in Table \ref{tab:Table 2}: For the purpose of this analysis, we classified all windows that were at least 20\% homozygous as ROH, and the remainder as non-ROH. Likewise, we classified all windows with a posterior probability of ROH of at least 20\% as ROH. We used these cases to compute sensitivity (Se) and specificity (Sp). In the control case (simulations with no ascertainment, contamination, or ROH) we see that the specificity remains $\geq0.99$ for all coverages. In the cases where we simulate ROH, we find that sensitivity decreases from 97\% at 4x coverage to less than 65\% at 0.05x coverage, but specificity remains high at above 0.96 for all cases, suggesting that the number of erroneously called ROH segments is low.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_review.png}
    \centering
    \caption{Estimation of ROH probabilities along the genome in simulations. The top row in each panel shows the proportion of differences in a simulated individual along the genome. In the bottom row, the dotted line shows the proportion of each window in ROH, and the solid lines show the estimated probability of not being in the ROH state. (A) Simulation with no ascertainment, contamination, or ROH. (B) Simulation with ROH. (C) Simulation with ROH and ascertainment. (D) Simulation with ROH and contamination.}
    \label{fig2:ROH}
\end{figure}

\begin{table}[h!]
\caption{\label{tab:Table 2}Model performance for ROH prediction. Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
    \hline
    Coverage & Control (Sp) & R (Se) & R (Sp) & RA (Se) & RA (Sp) & RC (Se) & RC (Sp)\\
    \hline
    4x & $>$0.99 & 0.97 & 0.98 & 0.97 & 0.98 & 0.97 & 0.97\\
    \hline
    0.5x & $>$0.99 & 0.96 & 0.98 & 0.96 & 0.98 & 0.96 & 0.97\\
    \hline
    0.2x & $>$0.99 & 0.92 & 0.98 & 0.91 & 0.98 & 0.93 & 0.97\\
    \hline
    0.1x & $>$0.99 & 0.85 & 0.98 & 0.83 & 0.97 & 0.82 & 0.97\\
    \hline
    0.05x & 0.99 & 0.65 & 0.96 & 0.55 & 0.96 & 0.57 & 0.96\\
    \hline
\end{tabular}
\label{table2}
\end{table}

\paragraph{IBD prediction:}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatedness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.78, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.48 at 0.03x. The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage. We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction. Adding ROH to the simulations reduces the IBD prediction in two cases: the average accuracy for second degree decreases from 0.89 to 0.85 and for siblings from 0.81 to 0.70 (Fig. S9). We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}). Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$). However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see Fig. \ref{fig3:Comparison_READ_KIN})

@@

\paragraph{Relatedness classification:}
We evaluated the classification accuracy of KIN (cutoff: 1 log-likelihood unit) and compared it to that of READ (cutoff: 1 standard deviation)) (Fig. \ref{fig3:Comparison_READ_KIN}). We first describe the results for relatedness cases detectable by READ, viz. identical, $1^{st}$ degree, $2^{nd}$ degree and unrelated. In this case we show that both methods have similar performance for low-coverage shotgun data (``control''-case), and for ascertained data. The true positive rate is above 0.97 for both KIN and READ, while the false negative rate is below 0.02. One exception is  $2^{nd}$-degree relatives, where KIN has higher power than READ, and as coverage decreases from 4x to 0.05x, KIN's true positive rate decreases from 0.95 to 0.89, compared to range of 0.91 to 0.75 for READ. The false positive rate in this case remains below 0.03 for both methods.

To investigate the impact of contamination, we performed simulations where we added up to 3\% contamination to some individuals. We also find that READ is strongly impacted by contamination  as the true positive rate is in the range 0.89 to 0.78 and 0.54 to 0.46 for 1st and 2nd degree relatives, respectively, and the false positive rate reaches up to 0.04 and 0.17 for unrelated individuals and 2nd degree relatives, respectively. In comparison, we also ran KIN, giving it the simulated contamination amounts for each individual. We found that the correction implemented in KIN is sufficient to remove this bias, and the true and false positive rates remain as in the control (Fig. \ref{fig3:Comparison_READ_KIN}). We also perform an analysis where we misspecify the contamination, and find that the effect of modest misspecification is small (Fig. S13).

For simulations with ROH, we find that KIN also outperforms READ for 1st and 2nd degree relatives, although both the true and false positive rates increased for both methods compared to the control for 2nd degree relatives. The increase in false positives is likely due to ROH making related individuals more similar. Finally, we also found that KIN has good power to detect relatedness cases that are not detectable by READ, i.e. parent-child, siblings and, to a lower level, 3rd degree relatives (Fig. \ref{fig3:Comparison_READ_KIN}).

@@

Only for 28 pairs, KIN has confident classifications that differ from those obtained using READ (table S2, Fig. S11). Twenty of these comparisons, KIN predicts to be third-degree relationships, and  READ concordantly classifies them as unrelated (as it only infers first and second degree relationships). lcMLkin predicts $3^{rd}$ - $5^{th}$ in 14 of these cases, unrelated in 1 case (see Fig. S11), second degree in 1 case (see Fig. S11), and does not have enough data for classification in 4 cases. For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^{th}$ degree, but KIN infers them to be unrelated. There are a total of 83 pairs where READ obtains a confident call ($s.d.>1$), but KIN does not ($\Delta LL<1$). In 80 of these cases, READ classifies them as unrelated, KIN classifies $3^{rd}$ degree, while lcMLkin predicts $3^{rd}$ - $5^{th}$ degree. The remaining cases READ classifies unrelated, while KIN and lcMLkin call a $2^{nd}$-degree relationship. Thus, the vast majority of differences can be explained by READ not considering 3rd-degree relationships.

@@

For detecting ROH, there is only one method available that works with low coverage ancient DNA \cite{ringbauer_parental_2021}. This software can infer ROH at coverages $\geq0.3x$, but requires a large reference panel. Instead, our method detects long ROH based on just the expected heterozygosity $p_0$, which can be estimated from a small number of unrelated individuals.  On simulations, we find that our method reliably inferrs ROH regions  with lengths on the order of 10cM from samples sequenced to coverage $\geq 0.1x$ . 
In our simulations, we find that adding long ROH ($\approx17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages. This is likely because ROH actually reduces the variance in differences depending on the relatedness, and makes it easier to correctly classify relatedness. For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias (see Fig. \ref{fig3:Comparison_READ_KIN}).

We show that KIN reliably detects IBD tracts for $\geq0.1x$ coverage even in the presence of contamination and ascertainment bias, and the accuracy for IBD detection reduces with lower coverages. Adding ROH to the simulations adversely affects IBD prediction in siblings (Fig. S9), but this does not affect the power to correctly classify siblings (Fig. \ref{fig3:Comparison_READ_KIN}).

@@

The Lech Valley data has low contamination, and no ROH. For pairwise comparisons with large numbers of overlapping sites ($>10000$), KIN, READ and lcMLkin all mostly agree. However, KIN is able to differentiate between parent-child and siblings, and identify second degree relationship from just a few thousand polymorphic sites ($\approx$ 4000) overlapping between samples. KIN can also infer third degree relation with $\approx$ 30,000 overlapping polymorphic sites. We show that when applied to Neandertal specimens from Chagyrskaya and Okladnikov Caves, KIN identifies a pair of $1^{st}$-Degree relatives as parent-child, which is in agreement with the finding that the mtDNA haplotypes differ between the samples (one sample is male and the other female) \cite{laurits_skov_genetic_nodate}. In addition, KIN identifies a pair of $3^{rd}$ degree relatives. In this case of a population with large amounts of ROH, we find that the inference by lcMLkin are heavily biased, but KIN's model takes ROH into account and both the coefficient of relatedness and $k_0$ are very close to what would be expected from the inference by both READ and KIN.
